Classification Techniques for Conformance and Performance checking in Process Analysis

### 2013-09-17 paper/paper2.tex
semplificata tutta la performance

\section{Introduction}

Today, many Information Systems that integrate the concept of business process record all events occurring during the process execution in an event log. An event log represents a trace of the process behavior that can be observed and analyzed in order to tune it with the business objectives pursued. Process Mining provides a set of techniques for process discovery and analysis. Through the knowledge extracted from event logs, process discovery allows the construction of a process model. Process analysis instead assumes the existence of a model and it consists of checking the conformance and performance of the process executions with respect to it.

@@

Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm of \cite{2}, and mapping each activity to a pair of start/end transitions. The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g. the \emph{Join} or \emph{Fork} gateway \cite{2}.

\section{Process Analysis}\label{Background}

Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are events. An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}. Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process are grouped into {\itshape traces}. Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions. In this paper the following assumptions are made:
\begin{itemize}
\item All traces are instances of the same process.
\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
\end{itemize}

\subsection{Log replay algorithm}\label{logreplayAlg}

The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}. Given a Petri net model and an event log as input to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics. For each trace in the log, the algorithm starts by placing one token in the start place of the net. For each event in the trace the corresponding transition is fired assuming a {\itshape non-blocking} behavior of the algorithm, then the marking of the net is updated. Non-blocking replay means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either creating artificial tokens in the pre-set or, if possible, firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function. The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.

@@

The log replay of the trace $T'$ terminates with remaining tokens \{${p5 \rightarrow 1,p13 \rightarrow 1}$\} and one missing token \{${p8 \rightarrow 1}$\}. The missing token is created artificially and this fact witness a wrong execution of the event $Notify\_start$. In fact, in $T'$ this event is executed before the termination of the activity \emph{FinancialCheck}: this is interpreted as a non-conformance to the process model.

@@

\begin{comment}
To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section \ref{example} and the trace T presented in Figure \ref{ConfLog}. The replay starts with firing the transition $Order\_start$ at time $0s$, so a token arrives at $p2$ at time $0s$. After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=45min$, and a token arrives at $p3$ and $p4$. At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$. At the time $1h:10min:20s$ the transition $FinancialCheck\_start$ is fired and a token is produced at $p5$, so $tw(p3)=25min+20s$. After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p8$. At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and a token is produced at $p7$. After $15 min$ the invisible transition $t1$ is fired and a token is produced at $p10$. At this point the transition $t2$ is enabled, so its firing is now possible and a token is produced at $p11$. Notice that $tsc(p8)=40min, tsc(p10)=0s$. At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.
\end{comment}

@@

@@

Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{5}. A \emph{Training set} consisting of records whose class labels are known must be provided to a learning algorithm, and it is used to build a classification model. This model is then applied to a \emph{test set}, which consists of records with unknown class labels. This application allows the estimation of the \emph{model accuracy} that can be done in different ways. The simple one consists in computing the \emph{performance} as the number of the records correctly predicted over the total number of records in the test set.

Many classification models could be used for a classification problem. For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly. This fact is fundamental for giving an explanation to the process behavior based on the process data values. Furthermore, this choice is encouraged by the simplicity of decision trees. In a decision tree, a class label (possible value of the attribute target) is assigned to each leaf node. The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
\begin{comment}
In principle, there are exponentially many decision trees that can be constructed from a given set of attributes. While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space. However, efficient algoritms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time. These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partioning the data. For instance Hunt's algorithm is one of them and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.
\end{comment}
\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a classification problem can be formulated. The classification dataset is extracted from the process data: for each process instance, one will extract some data that will be included as a record in the dataset. For the process presented in Section \ref{example}, a record is characterized by the following attribute set: an instance identifier, a client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued. The conformance result for the process instance provides an additional and important data which takes the role of the target attribute. All these attributes are discrete and contribute to build the dataset presented in table \ref{tab:SaleData} of the classification problem. Note that, since the goal of this paper is just to illustrate an approach for exploiting process data in the process analysis, the data presented here is just a synthetic data that was generated with some noise to emulate a real situation.

@@

Performance analysis of a process can be performed exploiting log replay. Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as the \emph{synchronization time} (i.e. time interval between arrival of a token in the place and enabling of a transition in the post-set of the place).

Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance. In order to understand the \emph{cause} of performance anomalies that can affect the process behavior, we have to explore the process data. Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services. For example, discovering data patterns in correspondence to which some activities need more time for completion than others, helps in making decisions about resources distribution to the process activities or in scheduling activities. In addition to the completion time, one could analyze a process under more complex performance metrics such as the \emph{synchronization time}. A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process. To find out the possible data influences on synchronization time, the approach presented in Section \ref{ClassConformance} for conformance analysis through classification can be easily extended to performance analysis. For example, a classification problem can be formulated for each synchronization point of the process. In this way, the model obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.

@@

To formulate the classification problem the dataset is extracted from the event log, whereas the target attribute is one of the log replay performance results. The target must be something that characterizes the performance in term of synchronization. An idea can be to simply consider the difference between synchronization time of $p8$ and $p10$: $tsc(p8) - tsc(p10)$. It should be noted that classification deals with discrete or categorical target attribute. Since the performance metrics, the synchronization time in this case, are expressed as continous values, a discretization phase is required  during the formulation of the classification problem. Let us choose the attribute target as a categorical attribute: given a treshold $S$, if $|tsc(p8)- tsc(p10)|$ exceeds the value $S$, then the target attribute indicates the branch with the high synchronization time, otherwise the target indicates that the instance presents a normal synchronization time. Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}. Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in table \ref{tab:SaleDataPerf}. As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure \ref{salesPerfDecTree}.

@@

\section{Implementation with ProM}\label{implementation}
In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6} \cite{6}, an open-source framework implementing Process Mining tools, and \emph{Weka} \cite{7}, a data mining framework providing machine learning tools.
We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
\begin{itemize}
\item \texttt{Generate Instances With Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay. It returns as output a training set as shown in Table \ref{tab:SaleData} in the format needed by Weka classification tool. Note that the plug-in extracts \emph{all} the attributes data recorded in the event log. In case, the dataset resulting can be subject of some \emph{feature selection} techniques before using it as a training set for the classification task.

\begin{comment}
\item \texttt{Generate Instances With Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay. The output is a training set as presented in table \ref{tab:SaleDataPerf}.
\end{comment}

\item \texttt{Generate Instances to Classify}: given an event log, the plug-in generates a dataset of instances that can be classified using a classification model.
\end{itemize}
The second category includes plug-ins for the classification model generation and utilization for the prediction purpose:
\begin{itemize}
\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using $J48$ algorithm, an implementation of the algorithm $C4.5$ developed by Weka.
\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classify the instances according to the classifier.
\end{itemize}
Finally, the last category includes a set of plug-ins useful for the serialization, deserialization and visualization of the resources used by the previous plug-ins, such as dataset and classifiers.

>>>>

\section{Introduction}
Today, many Information Systems that integrate the concept of business process record all events occurring during the process execution in an event log. An event log represents a trace of the process behavior that can be observed and analyzed in order to tune it with the business objectives pursued. Process Mining provides a set of techniques for process discovery and analysis. Through the knowledge extracted from event logs, process discovery allows the construction of a process model. Process analysis instead assumes the existence of a model and it consists of checking the conformance and performance of the process executions with respect to it.

@@

Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{2}, and mapping each activity to a pair of start/end transitions. The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g. the \emph{Join} or \emph{Fork} gateway \cite{2}.

\section{Process Analysis}\label{Background}
%Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are events. An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}. Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process are grouped into {\itshape traces}. Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are events. An event $e$ can be seen as a pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.  Events that belong to the same process are grouped into {\itshape traces}. A trace $T$ is a finite sequence of events ordered based on timestamp. A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.  In this paper the following assumptions are made:
\begin{itemize}
\item All traces are instances of the same process.
\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
\end{itemize}

\subsection{Log replay algorithm}\label{logreplayAlg}
The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}. Given a Petri net model and an event log as input to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics. For each trace in the log, the algorithm starts by placing one token in the start place of the net. For each event in the trace the corresponding transition is fired assuming a {\itshape non-blocking} behavior of the algorithm, then the marking of the net is updated. Non-blocking replay means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either creating artificial tokens in the pre-set or, if possible, firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function. %The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.

@@

The log replay of the trace $T'$ terminates with remaining tokens \{$p5 \rightarrow 1$, $p13 \rightarrow 1$\} and one missing token \{${p8 \rightarrow 1}$\}. The missing token is created artificially and this fact witness a wrong execution of the event $Notify\_start$. In fact, in $T'$ this event is executed before the termination of the activity \emph{FinancialCheck}: this is interpreted as a non-conformance to the process model.

@@

\begin{comment}
To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section \ref{example} and the trace T presented in Figure \ref{ConfLog}. The replay starts with firing the transition $Order\_start$ at time $0s$, so a token arrives at $p2$ at time $0s$. After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=45min$, and a token arrives at $p3$ and $p4$. At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$. At the time $1h:10min:20s$ the transition $FinancialCheck\_start$ is fired and a token is produced at $p5$, so $tw(p3)=25min+20s$. After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p8$. At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and a token is produced at $p7$. After $15 min$ the invisible transition $t1$ is fired and a token is produced at $p10$. At this point the transition $t2$ is enabled, so its firing is now possible and a token is produced at $p11$. Notice that $tsc(p8)=40min, tsc(p10)=0s$. At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.

@@

\end{comment}

@@

Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{5}. A \emph{Training set} consisting of records whose class labels are known must be provided to a learning algorithm, and it is used to build a classification model. This model is then applied to a \emph{test set}, which consists of records with unknown class labels. This application allows the estimation of the \emph{model accuracy} that can be done in different ways. The simple one consists in computing the accuracy as the number of the records correctly predicted over the total number of records in the test set.

Many classification models could be used for a classification problem. For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly. This fact is fundamental for giving an explanation to the process behavior based on the process data values. Furthermore, this choice is encouraged by the simplicity of decision trees. In a decision tree, a class label (possible value of the attribute target) is assigned to each leaf node. The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
\begin{comment}
In principle, there are exponentially many decision trees that can be constructed from a given set of attributes. While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space. However, efficient algoritms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time. These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partioning the data. For instance Hunt's algorithm is one of them and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.
\end{comment}
\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a classification problem can be formulated. The classification dataset is extracted from the process data: for each process instance, one will extract some data that will be included as a record in the dataset. For the process presented in Section \ref{example}, a record is characterized by the following attribute set: an instance identifier, a client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued. The conformance result for the process instance provides an additional and important data which takes the role of the target attribute. All these attributes are discrete and contribute to build the dataset presented in Table \ref{tab:SaleData} of the classification problem. Note that, since the goal of this paper is just to illustrate an approach for exploiting process data in the process analysis, the data presented here is just a synthetic data that was generated with some noise to emulate a real situation.

@@

Performance analysis of a process can be carried exploiting log replay. Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as: \emph{synchronization time} (i.e. time interval between arrival of a token in the place and enabling of a transition in the post-set of the place), \emph{sejour time} (i.e. the time interval between arrival and departure of tokens) and \emph{waiting time} (i.e. the time interval between enabling of a transition in the post-set of the place and token departure).

Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance. In order to understand the \emph{causes} of performance anomalies that can affect the process behavior, one can explore the process data. Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services. For example, discovering data patterns in correspondence to which some activities need more time for completion than others, helps in making decisions about resources distribution to the process activities or in scheduling activities. In addition to the completion time, one could analyze a process under more complex performance metrics such as the synchronization time. A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process. To find out the possible data influences on synchronization time, the approach presented in Section \ref{ClassConformance} for conformance analysis through classification can be easily extended to performance analysis. For example, a classification problem can be formulated for each synchronization point of the process. In this way, the classifier obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.

@@

To formulate the classification problem the dataset is extracted from the event log, whereas the target attribute is one of the log replay performance results. The target must be something that characterizes the performance in term of synchronization. An idea can be to simply consider the difference between synchronization time of $p8$ and $p10$: $tsc(p8) - tsc(p10)$. It should be noted that classification deals with discrete or categorical target attribute. Since the performance metrics, the synchronization time in this case, are expressed as continous values, a discretization phase is required  during the formulation of the classification problem. Let us choose the attribute target as a categorical attribute: given a treshold $S$, if $|tsc(p8)- tsc(p10)|$ exceeds the value $S$, then the target attribute indicates the branch with the high synchronization time, otherwise the target indicates that the instance presents a normal synchronization time. Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}. Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in Table \ref{tab:SaleDataPerf}. As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure \ref{salesPerfDecTree}.

@@

\section{Implementation with ProM}\label{implementation}
In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6} \cite{6}, an open-source framework implementing Process Mining tools, and \emph{Weka} \cite{7}, a data mining framework providing machine learning tools.
We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
\begin{itemize}
\item \texttt{Generate Instances With Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay. It returns as output a training set as shown in Table \ref{tab:SaleData} in the format needed by Weka classification tool. Note that the plug-in extracts \emph{all} the attributes data recorded in the event log. In case, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.

\begin{comment}
\item \texttt{Generate Instances With Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay. The output is a training set as presented in Table \ref{tab:SaleDataPerf}.
\end{comment}

\item \texttt{Generate Instances to Classify}: given an event log, the plug-in generates a dataset of instances that can be classified using a classification model.
\end{itemize}
The second category includes plug-ins for the classifier generation and utilization as a predictive model:
\begin{itemize}
\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using $J48$ algorithm, an implementation of the algorithm $C4.5$ developed by Weka.
\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classify the instances according to the classifier.
\end{itemize}
Finally, the last category includes a set of plug-ins useful for the serialization, deserialization and visualization of the resources used by the previous plug-ins, such as dataset and classifiers.
